---
title: "ECON 424 Chapter 1 Computer Exercises"
author: John Doe
date: "Monday, September 28, 2026"
output: word_document
---

*Before you knit, replace "John Doe" in the header above with your own name. Type your answers after each* **Answer:** *line.*

*Exercise and part numbers follow Wooldridge's Chapter 1 computer exercises; some are skipped.*

```{r message=F, warning=F, error=F}
library(wooldridge)
library(dplyr)
```

## C1

Use the data in WAGE1 for this exercise.

### (i) Find the average education level in the sample. What are the lowest and highest years of education?

```{r}
summary( wage1$educ )
```

**Answer:** The average education level in the sample is `r round( mean( wage1$educ ), 2 )` years. The lowest is `r min( wage1$educ )` years and the highest is `r max( wage1$educ )` years.

### (ii) Find the average hourly wage in the sample. Does it seem high or low?

**Answer:** The average hourly wage in the sample is \$`r sprintf( "%.2f", mean( wage1$wage ) )` per hour. *Your turn: does this seem high or low? Why?*

### (iii) The wage data are reported in 1976 dollars. Obtain and report the CPI for the years 1976 and 2013.

**Answer:** CPI for 1976 was 56.933. CPI for 2013 was 232.957.

### (iv) Use the CPI values from part (iii) to find the average hourly wage in 2013 dollars. Now, does the average hourly wage seem reasonable?

**Answer:** In 2013 dollars, the average is \$`r sprintf( "%.2f", mean( wage1$wage ) * (232.957 / 56.933) )` per hour. At 40 hours a week for 50 weeks, that's about \$48,000 a year, which seems reasonable.

### (v) How many women are in the sample? How many men?

There are a few ways of doing this. Here are two.

*Add up the number of "1's" in data\$female. Then subtract from n_obs to get the number of males.*

```{r}
sum( wage1$female )
length( wage1$female ) - sum( wage1$female )
```

*Use the dplyr package, and then the count command.*

```{r}
wage1 |> count(female == 1)
```

The \|\> command is called the pipe operator. It takes the data on the left, and plugs it into the function on the right.

Anything in a code chunk, or in a script file, that follows one or more of the number/pound symbol (\#) is a "comment" and isn't run. As you get further along, it is a VERY good idea to leave yourself notes through comments in your code.

You don't have to, but in the case of R Markdown it's a good idea to take out your comments before you compile your document, otherwise they show up in your document. You can always delete them after you knit, though.

**Answer:** There are `r sum( wage1$female )` women and `r length( wage1$female ) - sum( wage1$female )` men in the sample.

------------------------------------------------------------------------

## C2 Use the data in BWGHT to answer this question.

### (i) How many women are in the sample, and how many report smoking during pregnancy?

For the first part, all the observations in the data are of women. You can read the number in the "environment" tab, or you can calculate

```{r}
length( bwght$cigs )
```

For the second part, a few options. Easiest one is to use the dplyr library, and count command.

```{r}
bwght |> count( cigs > 0 )
```

**Answer:**

### (ii) What is the average number of cigarettes smoked per day? Is the average a good measure of the "typical" woman in this case? Explain.

Here you can use the mean() command on variable bwght\$cigs.

```{r}
mean ( bwght$cigs )
```

**Answer:**

### (iii) Among women who smoked during pregnancy, what is the average number of cigarettes smoked per day? How does this compare with your answer from part (ii), and why?

A few ways to do this as well. One is to only select observations for which cigs \> 0. In R you can use conditional statements in row/column selection. In this case, you want the mean of rows where cigs \> 0, and you want the mean of the "cigs" column.

```{r}
mean( bwght [ bwght$cigs > 0, "cigs" ] )
```

There's also a subset command. You can take a subset of your data -- I'm calling the subset "smoker" -- and then calculate the mean of the "cigs" values for the individuals in that subset.

```{r}
smoker <- subset( bwght, cigs > 0 )
mean( smoker$cigs )
```

**Answer:**

### (iv) Find the average of *fatheduc* in the sample. Why are only 1,192 observations used to compute this average?

There are quite a few NA's (not available) in this series, so you can't just use the mean() command. However, summary() works. So does using mean() with the option na.rm = TRUE, which tells R to remove the NA's before calculating.

```{r}
summary( bwght$fatheduc )

mean( bwght$fatheduc, na.rm = TRUE )
```

**Answer:**

### (v) Report the average family income and its standard deviation in dollars.

You can use the mean() command. And there's a sd() command for standard deviation. Be careful: *faminc* is measured in thousands of dollars, so you'll need to multiply by 1,000 to report your answers in dollars.

```{r}
mean( bwght$faminc ) * 1000
```

Now it's your turn. In the chunk below, copy the line above and change mean to sd.

```{r}
# Your turn: copy the line above and change mean to sd
<<<<<<< HEAD
sd( bwght$faminc ) * 1000
```

**Answer:** The standard deviation for the family income is 18739.28
=======
sd(bwght$faminc) * 1000
```

**Answer:** The standard deviation for the average family income is 18739.28

>>>>>>> 9559b91d83bf454abc1c24c16b801a887be37c24
------------------------------------------------------------------------

## C3 The data in MEAP01 are for the state of Michigan in the year 2001. Use these data to answer the following questions.

### (i) Find the largest and smallest values of *math4*. Does the range make sense? Explain.

You can use the summary() command.

```{r}
summary( meap01$math4 )
```

**Answer:** The largest number in math4 is 100 and the smallest is 0. This makes sense as this is usually the maximum and minimum scores on a test

### (ii) How many schools have a perfect pass rate on the math test? What percentage is this of the total sample?

Easiest way is the count() command. Then you can calculate the percentage: the number of schools with a perfect pass rate, divided by the total number of schools, times 100.

```{r}
meap01 |> count( math4 == 100 )

sum( meap01$math4 == 100 ) / length( meap01$math4 ) * 100
```

<<<<<<< HEAD

**Answer:** The number of schools with a perfect pass rate is 38. This only accounts for 2.08 percent of all schools in the dataset.
=======
**Answer:**
>>>>>>> 9559b91d83bf454abc1c24c16b801a887be37c24

### (iv) Compare the average pass rates for the math and reading scores. Which test is harder to pass?

```{r}
mean( meap01$math4 )
```

Now it's your turn. In the chunk below, copy the line above and change math4 to read4.

```{r}
# Your turn: copy the line above and change math4 to read4
<<<<<<< HEAD
mean( meap01$read4 )

```

**Answer:** According to the average test scores for both, the reading test is harder to pass
=======
mean ( meap01$read4 )
```

**Answer:** The 
>>>>>>> 9559b91d83bf454abc1c24c16b801a887be37c24

### (v) Find the correlation between *math4* and *read4*. What do you conclude?

Correlation command is cor().

```{r}
cor( meap01$math4, meap01$read4 )
```

**Answer:** The correlation of the math and reading scores is 0.84, meaning that they are strongly correlated and those who do well on one of them is likely to do well on the other test.

### (vii) Suppose School A spends \$6,000 per student and School B spends \$5,500 per student. By what percentage does School A's spending exceed School B's? Compare this to 100[log(6,000)-log(5,500)], which is the approximation percentage difference based on the difference in the natural logs.

The first line below calculates the exact percentage difference. In R the command for natural log is log(), so the second line calculates the log approximation.

```{r}
( 6000 - 5500 ) / 5500 * 100

100 * ( log(6000) - log(5500) )
```

**Answer:** The spending for school A exceeds the spending for school B by 9.09%. This is close to the log approximation, which is 8.70%

------------------------------------------------------------------------

## C4 The data in JTRAIN2 come from a job training experiment conducted for low-income men during 1976-1977; see Lalonde (1986).

### (i) Use the indicator variable *train* to determine the fraction of men receiving job training.

The variable *train* is binary (1 if the man received job training, 0 if not), so its average is the fraction of men who received training.

```{r}
mean( jtrain2$train )
```

**Answer:** The percentage of men that received training is 41 percent.

### (ii) The variable *re78* is earnings from 1978, measured in thousands of 1982 dollars. Find the averages of *re78* for the sample of men receiving job training and the sample not receiving job training. Is the difference economically large?

Here you'll subset the data again.

```{r}
trained <- subset( jtrain2, train > 0 )
not_trained <- subset(jtrain2, train == 0 )
```

Then you take the mean of *re78* for each group.

```{r}
mean( trained$re78 )

mean( not_trained$re78 )
```

<<<<<<< HEAD

**Answer:** The average earnings of men in 1978 who receive job training is 6.35 while the average of earnings of men who do not receive job training is 4.55.
=======
**Answer:**
>>>>>>> 9559b91d83bf454abc1c24c16b801a887be37c24

### (iii) The variable *unem78* is an indicator of whether a man is unemployed or not in 1978. What fraction of the men who received job training are unemployed? What about for men who did not receive job training? Comment on the difference.

Use the trained and not_trained subsets from before.

```{r}
mean( trained$unem78 )
```

Now it's your turn. In the chunk below, copy the line above and change trained to not_trained.

```{r}
# Your turn: copy the line above and change trained to not_trained
mean( not_trained$unem78)

```

**Answer:** The fraction of men that receive job training that are unemployed is 0.24 while th e fraction of men that do not receive job training and that are unemployed is 0.35

### (iv) From parts (ii) and (iii), does it appear that the job training program was effective? What would make our conclusions more convincing?

**Answer:** From the previous parts, it seems that job training does have a correlation when it comes to better average salaries and better employment rates. Our studies would be more convincing if we were able to find other variables that are related to better outcomes for people with job training. 

------------------------------------------------------------------------

## C6 Use the data in COUNTYMURDERS to answer this question. Use only the year 1996. The variable *murders* is the number of murders reported in the county. The variable *execs* is the number of executions that took place of people sentenced to death in the given county. Most states in the United States have the death penalty, but several do not.

### (i) How many counties are there in the data set? Of these, how many have zero murders? What percentage of counties have zero executions? (Remember, use only the 1996 data.)

First subset the data:

```{r}
countymurders96 <- countymurders |> subset( year == 1996 )
```

```{r}
length( countymurders96$countyid )

countymurders96 |> count( murders == 0 )

mean( countymurders96$execs == 0 ) * 100
```

The last line uses a handy trick: *execs == 0* is TRUE (1) or FALSE (0) for each county, and the average of a variable that only takes the values 0 and 1 is the proportion of 1's. So its mean is the proportion of counties with zero executions. Multiply by 100 to get a percentage.

**Answer:** The amount of counties in the 1996 data is 2197. Out of those, only 1051 had 0 murders. The percentage of counties that have 0 executions 98 percent of counties

### (ii) What is the largest number of murders? What is the largest number of executions? Why is the average number of executions so small?

```{r}
max( countymurders96$murders )

max( countymurders96$execs )

mean(countymurders96$execs)
```

**Answer:** The largest number of murders in a county in 1996 was 1403. The largest number of executions in a county in 1996 was 3. The average number of executions is so small since the majority of counties do not do executions, as we saw that 98 percent of counties have not had an execution from the last problem.

### (iii) Compute the correlation coefficient between *murders* and *execs* and describe what you find.

```{r}
cor( countymurders96$murders, countymurders96$execs )
```

**Answer:** The correlation between murders and executions is 0.210, which is low and implies that there is not a very strong relationship between murders and executions

### (iv) You should have computed a positive correlation in part (iii). Do you think that more executions *cause* more murders to occur? What might explain the positive correlation?

**Answer:** There seems to be a reverse causality here. More people committing murders would lead to more people getting the death penalty and driving up executions.


------------------------------------------------------------------------

## C8 The data in ECONMATH were obtained on students from a large university course in introductory microeconomics. For this problem, we are interested in two variables: *score*, which is the final course score, and *econhs*, which is a binary variable indicating whether a student took an economics course in high school.

### (i) How many students are in the sample? How many students report taking an economics course in high school?

```{r}
length(econmath$score)
sum(econmath$econhs)

```

**Answer:** The amount of students in the sample is 856. There are 317 people that reported taking an economics course in high school

### (ii) Find the average of *score* for those students who did take a high school economics class. How does it compare with the average of *score* for those who did not?

```{r}
with_econhs <- subset( econmath, econhs == 1 )
no_econhs <- subset( econmath, econhs == 0 )

mean( with_econhs$score )
mean( no_econhs$score )



```

**Answer:** The average score for someone that took Econ in high school is 72.08. The average score for the students whom did not take econ was 72.91.

### (iii) Do the findings in part (ii) necessarily tell you anything about the causal effect of taking high school economics on college course performance? Explain.

**Answer:** The findings from the previous part do not tell us anything about the causal effect of taking high school economics since the average scores are very similar. It is also not reliable since students choose to take economics in high school, which might suggest some interest in the course.We also do not factor in the students relative motivation when it comes to studying in general.

### (iv) If you want to obtain a good causal estimate of the effect of taking a high school economics course using the difference in averages, what experiment would you run?

**Answer:** The experiment I would run is to randomly assign a group of students to high school economics, and for the rest of the students I would not have them taking it. This is more reliable as we won't have to worry about people choosing economics on their own.


